---
title: "NHANES Analysis"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output: pdf_document
---

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(cmdstanr)
library(tidybayes)
library(patchwork)
library(knitr)
library(rstiefel)
library(parallel)
library(doParallel)
library(loo)
source("../utility_functions.R")
options(mc.cores = parallel::detectCores() - 1)
knitr::opts_chunk$set(echo = FALSE)
```

# List version and package information

```{r}
R.Version()$version.string

print("Package versions:")
installed.packages()[names(sessionInfo()$otherPkgs), "Version"] %>% kable()
```

```{r, message=FALSE, warning=FALSE}
df <- read_csv("data/nhanes_data.csv")


## ALQ121: 1 every day or 2 nearly every day a drink

## ``The one treated individual consumed between one and three drinks on most days.
## The controls drank once a week or less, and most did not drink at all.'' (Rosenbaum, 2021)
df %>%
  select(SEQN, ALQ121, ALQ130) %>%
  filter(!is.na(ALQ121) & !is.na(ALQ130)) %>%
  filter(ALQ121 %in% c(1, 2) | ALQ121 >= 5) %>%
  filter(ALQ130 < 3) %>%
  mutate(light_drinking = ALQ121 %in% c(1, 2)) ->
drinking_df

## Fraction of treated units
print(sprintf("Fraction of treated units is %f", mean(drinking_df$light_drinking)))

## Demographics data frame
## RIDAGEYR - Age in years at screening - 0 to 24 mos
## RIAGENDR - Gender: 1 male, 2 female
## DMDEDUC2 - Education 1: less than 9th grade, 5 college or above
drinking_SEQN <- drinking_df$SEQN
demo_df <- df %>%
  dplyr::select(SEQN, RIDAGEYR, RIAGENDR, DMDEDUC2) %>%
  filter(SEQN %in% drinking_SEQN)

## LBDBGMSI - Mercury, methyl (nmol/L)
## LBDHDDSI HDL cholesterol
outcome_df <- df %>% filter(SEQN %in% drinking_SEQN)

## Gather all relevant outcomes
outcomes <- outcome_df %>%
  mutate(
    BPXODI = (BPXODI1 + BPXODI2 + BPXODI3) / 3,
    BPXOSY = (BPXOSY1 + BPXOSY2 + BPXOSY3) / 3
  ) %>%
  dplyr::select(
    SEQN, LBDHDDSI, LBDBGMSI, LBXSGL,
    LBXSKSI, LBXSNASI, LBDSIRSI,
    LBDBPBSI, LBDLDNSI, LBDTRSI, LBDBCDSI
  ) %>%
  drop_na() %>%
  arrange(SEQN)

## gather predictors: age, gender and education
predictors <- left_join(demo_df, drinking_df) %>%
  filter(SEQN %in% outcomes$SEQN) %>%
  arrange(SEQN) %>%
  dplyr::select(SEQN, light_drinking, RIDAGEYR, RIAGENDR, DMDEDUC2) %>%
  drop_na()

X <- predictors %>%
  mutate(
    light_drinking = ifelse(light_drinking, 1, 0),
    age = RIDAGEYR,
    gender = ifelse(RIAGENDR == 2, 1, 0),
    educ = ifelse(DMDEDUC2 <= 3, 0, 1)
  ) %>%
  select(-one_of("SEQN", "RIDAGEYR", "RIAGENDR", "DMDEDUC2"))

print("Summary information about treatment and covariates")
summary(X) %>% kable(digits = 2)
```

\newpage
# Calibration

Compute propensity scores given observed covariates 

```{r, warning=FALSE}
## Calibration: fitted values for treatment given covariates

preds_full <- predict(
  glm(light_drinking ~ RIDAGEYR + RIAGENDR + (DMDEDUC2 <= 3),
    data = predictors, family = "binomial"
  ),
  type = "response"
)

preds_no_educ <- predict(
  glm(light_drinking ~ RIDAGEYR + RIAGENDR,
    data = predictors, family = "binomial"
  ),
  type = "response"
)

preds_no_age <- predict(glm(light_drinking ~ RIAGENDR + (DMDEDUC2 <= 3), data = predictors, family = "binomial"), type = "response")
preds_no_gender <- predict(glm(light_drinking ~ RIDAGEYR + (DMDEDUC2 <= 3), data = predictors, family = "binomial"), type = "response")

## Compute odds ratios with and without benchmark covariate
or_educ <- ((preds_full / (1 - preds_full)) / (preds_no_educ / (1 - preds_no_educ)))
or_age <- ((preds_full / (1 - preds_full)) / (preds_no_age / (1 - preds_no_age)))
or_gender <- ((preds_full / (1 - preds_full)) / (preds_no_gender / (1 - preds_no_gender)))
or_educ <- ((preds_full / (1 - preds_full)) / (preds_no_educ / (1 - preds_no_educ)))

or_all <- ((preds_full / (1 - preds_full)) / (mean(predictors$light_drinking) / (1 - mean(predictors$light_drinking))))

## Some exploratory plots
age_odds_plot <- tibble(age_or = or_age) %>%
  ggplot() +
  geom_histogram(aes(x = age_or)) +
  theme_bw() +
  ylab("Count") +
  xlab("Change in Odds Ratio") +
  ggtitle("Age")

gender_odds_plot <- tibble(gender_or = or_gender) %>%
  ggplot() +
  geom_histogram(aes(x = gender_or)) +
  theme_bw() +
  ylab("Count") +
  xlab("Change in Odds Ratio") +
  ggtitle("Gender")

educ_odds_plot <- tibble(educ_or = pmax(1 / or_educ, or_educ)) %>%
  ggplot() +
  geom_histogram(aes(x = educ_or)) +
  theme_bw() +
  ylab("Count") +
  xlab("Change in Odds Ratio") +
  ggtitle("Educ") +
  xlim(c(1, 2))

age_odds_plot + gender_odds_plot + educ_odds_plot

### R-squared
rsquared_full <- summary(lm(light_drinking ~ age + educ + gender, data = X))$r.squared
rsquared_no_age <- summary(lm(light_drinking ~ educ + gender, data = X))$r.squared
rsquared_no_gender <- summary(lm(light_drinking ~ age + educ, data = X))$r.squared
rsquared_no_educ <- summary(lm(light_drinking ~ age + gender, data = X))$r.squared

## Partial R-squareds
rsq_partial_no_age <- (rsquared_full - rsquared_no_age) / (1 - rsquared_no_age)
rsq_partial_no_gender <- (rsquared_full - rsquared_no_gender) / (1 - rsquared_no_gender)
rsq_partial_no_educ <- (rsquared_full - rsquared_no_educ) / (1 - rsquared_no_educ)

## Benchmarks
age_benchmark <- rsq_partial_no_age / (1 - rsq_partial_no_age)
gender_benchmark <- rsq_partial_no_gender / (1 - rsq_partial_no_gender)
educ_benchmark <- rsq_partial_no_educ / (1 - rsq_partial_no_educ)
```

### Compute benchmark Lambda
Compute benchmark lambda values by calibrating in R2 space and converting black 
(see paper Sections 5.1 and 5.2)
```{r}
## Benchmark in lambda notation
lambda_age <- r2_to_odds(age_benchmark, prob = preds_full)
lambda_gender <- r2_to_odds(gender_benchmark, prob = preds_full)
lambda_educ <- r2_to_odds(educ_benchmark, prob = preds_full)

cat(sprintf(
  "Benchmark Lambdas:\nLambda_age = %f\nLambda_gender = %f\nLambda_educ =%f",
  lambda_age, lambda_gender, lambda_educ
))
```

# Histograms of outcomes

```{r, message=FALSE}
## Treatment variable
tr <- X[, "light_drinking"]

## Outcome variables
Y <- outcomes %>%
  filter(SEQN %in% predictors$SEQN) %>%
  arrange(SEQN) %>%
  dplyr::select(-SEQN) %>%
  mutate_all(log) %>%
  as.matrix()
colnames(Y) <- c(
  "HDL", "Methylmercury", "Glucose",
  "Potassium", "Sodium", "Iron", "Lead", "LDL", "Triglycerides", "Cadmium"
)

## Set values at measurement threshold to NA
Y[Y[, "Methylmercury"] == log(0.9), "Methylmercury"] <- NA
Y[Y[, "Cadmium"] == log(0.632), "Cadmium"] <- NA
bounds_mat <- matrix(c(2, -Inf, min(Y[, "Methylmercury"], na.rm = TRUE), 10, -Inf, min(Y[, "Cadmium"], na.rm = TRUE)), nrow = 2, byrow = TRUE)

## Impute missing values using Amelia
Y <- Amelia::amelia(Y, bounds = bounds_mat, p2s = 0)$imp[[1]]

yres <- (sapply(1:10, function(x) lm(scale(Y[, x]) ~ as.matrix(X))$residuals))
colnames(yres) <- colnames(Y)

as_tibble(yres) %>%
  pivot_longer(cols = 1:10, values_to = "value", names_to = "outcome") %>%
  ggplot() +
  geom_histogram(aes(x = value)) +
  facet_wrap(~outcome) +
  theme_bw()

## Regress each outcome on all covariates and compute p-values
outcome_pvals <- sapply(colnames(Y), function(x) coef(summary(lm(as.formula(sprintf("%s ~ light_drinking + age + gender + educ", x)), cbind(Y, X))))[2:5, 4]) < 0.05

## And R^_{Y ~ X} given all covariates without benchmark
rsq_full <- sapply(colnames(Y), function(x) summary(lm(as.formula(sprintf("%s ~ light_drinking + age + gender + educ", x)), data = cbind(Y, X)))$r.squared)
rsq_no_age <- sapply(colnames(Y), function(x) summary(lm(as.formula(sprintf("%s ~ light_drinking + gender + educ", x)), data = cbind(Y, X)))$r.squared)
rsq_no_gender <- sapply(colnames(Y), function(x) summary(lm(as.formula(sprintf("%s ~ light_drinking + age + educ", x)), data = cbind(Y, X)))$r.squared)
rsq_no_educ <- sapply(colnames(Y), function(x) summary(lm(as.formula(sprintf("%s ~ light_drinking + age + gender", x)), data = cbind(Y, X)))$r.squared)

partial_gender <- (rsq_full - rsq_no_gender) / (1 - rsq_no_gender)
partial_age <- (rsq_full - rsq_no_age) / (1 - rsq_no_age)
partial_educ <- (rsq_full - rsq_no_educ) / (1 - rsq_no_educ)
```

\newpage
# Run stan model and process results

Note: chunk option is set to FALSE by default (don't rerun inference every time we knit).
Focus on rank 5 model for analysis

```{r, eval=FALSE}
sm <- cmdstanr::cmdstan_model("../stan/multivariate_regression.stan")

## This may take some time
totalCores <- detectCores()
# Leave one core to avoid overload your computer
cluster <- makeCluster(totalCores[1] - 1)
registerDoParallel(cluster)

foreach(m = 1:6) %dopar% {
  data_list <- list(K = ncol(Y), J = ncol(X), N = nrow(Y), M = m, x = X, y = scale(Y))
  stan_results <- sm$sample(data = data_list)
  stan_results$save_object(file = sprintf("drinking_samples_weduc_rank%i.RDS", m))
}
sm_full <- cmdstanr::cmdstan_model("../stan/multivariate_regression_full_cov.stan")
data_list <- list(K = ncol(Y), J = ncol(X), N = nrow(Y), x = X, y = scale(Y))
stan_results <- sm_full$sample(
  data = data_list, chains = 4,
  cores = 4, max_treedepth = 15, adapt_delta = 0.95
)
stan_results$save_object(file = "drinking_samples_weduc_full_rank.RDS")

stopCluster(cluster)
```


```{r}
########### Process Results for visualization

m <- 5

stan_results <- readRDS(file = sprintf("drinking_samples_weduc_rank%i.RDS", m))
data_list <- list(K = ncol(Y), J = ncol(X), N = nrow(Y), M = m, x = X, y = scale(Y))
NC_index <- 2

naive_estimates <- stan_results %>%
  spread_draws(beta[K, J]) %>%
  group_by(J, K) %>%
  summarize(
    q025 = quantile(beta, 0.025),
    q975 = quantile(beta, 0.975)
  ) %>%
  mutate(sig = sign(q025 * q975) == 1) %>%
  mutate(feature = colnames(Y)) %>%
  filter(J == 1) %>%
  arrange(desc(q025)) %>%
  relocate(feature, q025, q975)
```

## Posterior mean and intervals for causal effects under NUC
```{r}
naive_estimates %>% knitr::kable(digits = 3)
```

# Figure 2 in paper
```{r}
## Find r2 closest to odds
r2_benchmark <- age_benchmark

## No single outcome confounding
r2y <- rep(NA, 10)
NC_index <- 2

beta_w_norm <- get_beta_w_norm(stan_results, data_list,
  NC_index = NC_index,
  probs = preds_full, full_rank = FALSE,
  R2_no_nc = r2_benchmark, R2_nc = 0.0,
  r2y = r2y
)
effect_intervals <- compute_effect_intervals(beta_w_norm)

nc_plot_index <- which(effect_intervals$K %in% c(NC_index))
make_interval_plot(effect_intervals,
  nc_plot_index = nc_plot_index,
  title = "Factor Confounding",
  labels = c("NUC", expression(Lambda[0.95] ~ "=" ~ Lambda[min]))
) -> plt

plt +
  scale_color_manual(
    name = "",
    labels = c(
      "NUC",
      bquote(paste(Lambda[0.95], "=", .(round(lambda_age, 1)))),
      expression(Lambda[0.95] ~ "=" ~ Lambda[min])
    ),
    values = c(nuc = "black", rect = "black", "null" = "#CC1C2F")
  ) +
  theme(legend.key = element_rect(color = c(NA, "black", NA), fill = c(NA, NA, NA)))

ggsave(file = sprintf("effect_intervals_weduc_%s_odds_m%i_new.pdf", paste(NC_index, collapse = "_"), m), width = 10, height = 5)
```

# R2Y = 1 for Mercury (Appendix Figure 4)
```{r}
## R2y=1 for mercury
NC_index <- 2
r2y <- rep(NA, 10)
r2y[2] <- 1

beta_w_norm <- get_beta_w_norm(stan_results, data_list,
  NC_index = NC_index,
  probs = preds_full, full_rank = TRUE,
  R2_no_nc = r2_benchmark, R2_nc = 0.0,
  r2y = r2y
)
effect_intervals <- compute_effect_intervals(beta_w_norm)
nc_plot_index <- which(effect_intervals$K %in% c(NC_index))
make_interval_plot(effect_intervals,
  nc_plot_index = nc_plot_index,
  title = expression(R[Y ~ "~" ~ "U|X"]^2 ~ "=1 for methylmercury"),
  labels = c("NUC", expression(Lambda[0.95] ~ "=" ~ Lambda[min]))
) -> plt

plt +
  scale_color_manual(
    name = "", labels = c(
      "NUC",
      bquote(paste(Lambda[0.95], "=", .(round(lambda_age, 1)))),
      expression(Lambda[0.95] ~ "=" ~ Lambda[min])
    ),
    values = c(nuc = "black", rect = "black", "null" = "#CC1C2F")
  ) +
  theme(legend.key = element_rect(color = c(NA, "black", NA), fill = c(NA, NA, NA)))

ggsave(file = sprintf("effect_intervals_%s_odds_m%i_null_r2y1.pdf", paste(NC_index, collapse = "_"), m), width = 10, height = 5)
```



\newpage
# Partial coefficients of determination (Appendix, Table 1)
```{r}
r2y_nsoc <- apply(matrix(unlist(compute_r2y(stan_results)), nrow = 10), 1, median)

##
rsq_x_mat <- apply(round(cbind(partial_age, partial_gender, partial_educ), digits = 3), 2, as.character)
for (i in 1:3) {
  sig_indices <- which(!t(outcome_pvals)[, i + 1])
  rsq_x_mat[sig_indices, i] <- "NS"
}
rsq_x_mat <- cbind(rsq_x_mat, round(r2y_nsoc, 3))
rownames(rsq_x_mat) <- colnames(Y)
colnames(rsq_x_mat) <- c("R2_Y~Age|-Age", "R2_Y~Gender|-Gender", "R2_Y~Educ|-Educ", "R2_Y~U|X")

rsq_x_mat %>% knitr::kable(format = "pandoc", digits = 2)
```

\newpage
# Model checking: Table 2 from Appendix
```{r, message=FALSE}
library(loo)

## Get PSIS Loo results from each stan object
loo_results <- lapply(1:6, function(i) {
  model_res <- readRDS(file = sprintf("drinking_samples_weduc_rank%i.RDS", i))
  model_res$loo()
})
names(loo_results) <- paste0("rank", 1:6)

## Add full rank
model_res <- readRDS(file = sprintf("drinking_samples_weduc_full_rank.RDS"))
loo_results[["full_rank"]] <- model_res$loo()

## Compare
comp <- loo::loo_compare(x = loo_results)
comp[, 1:2] %>% knitr::kable(format = "pandoc", digits = 2)
```


\newpage
# Gamma heat map: Figure 3 (Appendix)
```{r}

## Compute posterior mean of gamma (gamma is "B" in stan)
stan_results %>%
  spread_draws(B[K, M], lambda[K]) %>%
  nest(data = c(B, K, M, lambda)) %>%
  {
    map(.$data, function(x) {
      B <- matrix(x$B, nrow = max(x$K), ncol = max(x$M), byrow = TRUE)
      lambda <- x %>%
        filter(M == 1) %>%
        pull(lambda) %>%
        diag()

      B %*% t(B)
    })
  } -> BBlist

eig <- eigen(Reduce("+", BBlist) / 4000)

summary(sapply(BBlist, function(x) cov2cor(x)[1, 2]))

gamma_cols <- data_list$M # data_list$K
gamma_init <- eig$vectors[, 1:gamma_cols] %*% sqrt(diag(eig$values[1:gamma_cols]))
gamma_init <- varimax(gamma_init)$loadings

## Some manual rotation of Gamma and reordering of rows for presentation

g2 <- gamma_init[2, ]
g2 <- g2 / sqrt(sum(g2^2))
n2 <- rstiefel::NullC(g2)

g1 <- n2 %*% t(n2) %*% gamma_init[1, ]
g1 <- g1 / sqrt(sum(g1^2))

n21 <- rstiefel::NullC(cbind(g2, g1))

g8 <- n21 %*% t(n21) %*% gamma_init[8, ]
g8 <- g8 / sqrt(sum(g8^2))

n218 <- rstiefel::NullC(cbind(g2, g1, g8))

g4 <- n218 %*% t(n218) %*% gamma_init[4, ]
g4 <- g4 / sqrt(sum(g4^2))

n2184 <- rstiefel::NullC(cbind(g2, g1, g8, g4))

g3 <- n2184 %*% t(n2184) %*% gamma_init[3, ]
g3 <- g3 / sqrt(sum(g3^2))
nrest <- rstiefel::NullC(cbind(g2, g1, g8, g4, g3))


gamma <- gamma_init %*% cbind(g2, g1, g8, g4, g3)

ord <- c(2, 1, 8, 4, 3, 6, 9, 7, 10, 5)
gamma <- gamma[ord, ]

## Make the plot
tibble(expand.grid(outcome = colnames(Y)[ord], col = 1:gamma_cols)) %>%
  mutate(Value = as.numeric(gamma)) %>%
  ggplot(aes(x = col, y = forcats::fct_rev(as.factor(outcome)))) +
  geom_tile(aes(fill = Value), col = "black") +
  geom_text(aes(label = round(Value, 2)), col = "black") +
  colorspace::scale_fill_continuous_diverging() +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("") +
  ylab("") +
  ggtitle(expression(Gamma)) +
  theme(text = element_text(size = 20)) -> gamma_fig

gamma_fig

ggsave(gamma_fig, filename = sprintf("nhanes_heat_rank_%i.pdf", data_list$M), width = 6, height = 5)
```

